---
title: TD ecosystem functions class
author: Nina Schiettekatte
date: '2020-12-09'
slug: td-ecosystem-functions-class
categories: []
tags: []
---

This exercise consists of two parts:    
1) Fitting a Von Bertalanffy growth curve     
2) Quantifying functions for fish assemblages    

## Von Bertalanffy growth curve

This growth model is the most commonly used for fishes and can be fitted to length-age data (with the use of otoliths).  
It uses the following formula:  
$$
L = L_{inf} (1 - exp(-k(t-t_0))),
$$
where $L$ is the predicted length, $t$ is the age, $L_{inf}$ is the asymptotic average length, $K$ is the growth rate coefficient, $t_0$ is a modeling parameter represents the time or age where the average length is zero.     
<br>
We can try to fit this model to extract the three parameters. To do this, we will need to use the R function `nls()` (non-linear least squares regression).   
<be>
The syntax for creating a nonlinear least square test in R is:       
`nls(formula, data, start)`       

  - formula is a nonlinear model formula including variables and parameters (In our case the Von Bertalanffy model).

  - data is a data frame used to evaluate the variables in the formula.

  - start is a named list or named numeric vector of starting estimates (**TIP**: use `list(linf = 20, k = 0.5, t0 = 0)`).    
  
<br>
First things first, let's simulate some data to use. Copy the code block below and simply run it in R to simulate data.     

```{r}
library(dplyr)
data <- data.frame(
  age = as.numeric(rpois(100, 5)),
  linf = rnorm(100, 22, 2),
  k = rnorm(100, 0.7, 0.1)
) %>%
  mutate(length = linf * (1 - exp(-k*age))) %>%
  select(age, length)
```

We now created a dataframe with 100 observations of length and age of a fish species. We will use this data to fit our model. First we have a quick look at the data. 
```{r}
library(ggplot2)
ggplot(data) +
  geom_point(aes(x = age, y = length)) +
  theme_bw()
```


Go ahead and try to fit the model with the `nls()` function.      
**TIP**: use `start = list(linf = 20, k = 0.5, t0 = 0)`.


```{r}
fit <- nls(length ~ linf * (1 - exp(-k*(age - t0))), 
           data = data, 
           start = list(linf = 20, k = 0.5, t0 = 0))

```
What are the estimated values for k, linf and t0?
```{r}
summary(fit)
```
Now add a column to our dataframe with the predictions and make a figure. (**TIP** If tou use ggplot2, use the geom_smooth() option)

```{r}
data <- data %>%
  mutate(l_predict = predict(fit))

ggplot(data) +
  geom_point(aes(x = age, y = length)) +
  geom_smooth(aes(x = age, y = l_predict), se = FALSE, color = "red") +
  theme_bw() +
  labs(x = "Age (years)", y = "Length (cm)")

```




